---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

```{r load-data}
load("brfss2013.RData")
dim(brfss2013)
```
The data consist of 491775 observations (rows) and 330 variables (columns).


## Part 1: Data

BRFSS - Introduction

The CDC website states that, "The Behavioral Risk Factor Surveillance System (BRFSS) is the nation's premier system of health-related telephone surveys that collect state data about U.S. residents regarding their health-related risk behaviors, chronic health conditions, and use of preventive services. Established in 1984 with 15 states, BRFSS now collects data in all 50 states as well as the District of Columbia and three U.S. territories. BRFSS completes more than 400,000 adult interviews each year, making it the largest continuously conducted health survey system in the world."

Survey Methodology

The BRFSS survey collects data using a landline telephone and cellular telephones. The data collected via landline telephone survey is done by picking the respndents randomly who are adult from an household. In cellular telephone survey, the respondednts are not randomly selected but the adult people who volunteers are selected, also, the volunteers resides in private residence or college housing.

Implication of the data collection method on the scope of inference

Causality: Since the data collected is observational and no experiments are carried out, so we cannot define causation between related variables but we can associate their relationship. 

Generalizability: The data collected is huge and spreads across all the areas of the United States, so the data approximately captures the overall picture and can be generalized to US population.

There seem to be some bias in data due to underreporting or overreporting as the data is directly collected from the people and is not checked.

## Part 2: Research questions

**Research quesion 1:**

Among the people who use tobacco products daily (either smokeless or in smoke form) what is the effect of sleep hours on their health? 
This question will help CDC to figure out what are the sleep patterns of people who use tobacco products daily and how much this affect their health.
The variables used are:  comslept, smokday2, usenow3, X_rfhlth. The comslept is a variable created to divide the number of hours slept in 3 different categorical groups, other variables are used from those defined by BRFSS.

**Research quesion 2:**

For people who earns more than $75,000, what is the sleep pattern for those who are self-employed and those who are employed on wages?

This will help to know the amount of work load on people who are self-employed compared to those who are employed on wages. The variables used are: sleptim1, employ1, income2.

**Research quesion 3:**

What is the relation between the mental health of the people who earn more than $75,000 annually and those who earn less than $10,000 annually.

This will help to know the mental status of people who has high income and those who have poor income. The variables used are: menthlth and income2.


## Part 3: Exploratory data analysis






**Research quesion 1:**

```{r}
brfss2013 <- brfss2013%>%
  mutate(comslept =  ifelse(sleptim1 %in% 1:5,"sleep less than 6 hours", ifelse(sleptim1 %in% 6:8,"sleep between 6 to 8 hours",ifelse(sleptim1 %in% 9:24, "sleep more than 8 hours",NA))))
```
Here we added a new variable comslept to the brfss2013 data. This is a computed variable that divides number of hours slept per day into 3 groups: sleep less than 6 hours, sleep between 6 to 8 hours, sleep more than 8 hours. 

```{r}
tobacco <- brfss2013%>%
  select (comslept,smokday2,usenow3,X_rfhlth)%>%
  filter(!is.na(comslept) & (smokday2 == "Every day"|usenow3 == "Every day") & X_rfhlth != "NA") %>%
  group_by(X_rfhlth,comslept)%>%
  summarize(count = n())
tobacco
```

tobacco is a data frame which we created to analyse the research question. The data for the research question is summarized below. It shows the count of people who use tobacco products everyday and compares their sleep pattern with general health. 

```{r}
ggplot(tobacco, aes(x= comslept, y=count,fill = X_rfhlth)) + geom_col(position = "fill")                
```

This plot gives us the well defined and clear conclusions about the answer we were finding. The people who use tobacco products every day and sleep around 6 to 8 hours daily are found to be healthy than those who slept for more than 8 hopurs and less than 8 hours. People who slept less than 6 hours were found to have poor health due to less sleep. 







**Research quesion 2:**

```{r}
employ<- brfss2013%>%
  select(sleptim1,employ1,income2)%>%
  filter(income2 == "$75,000 or more" & employ1 %in% c("Employed for wages","Self-employed") & !sleptim1 %in% c("450","103") & sleptim1!= "NA")%>%
  group_by(sleptim1,employ1)%>%
  summarize(count=n())
employ
```

The data frame employ is created to store all the computations required for research question 2. The summary of the sleep time according to the type of employment was delineated in this reasearch question.

```{r}
ggplot(employ, aes(x=sleptim1,y=count,fill=employ1)) + geom_col(position="dodge")

```

The plot shows us the comparison of the amount of sleep that the people who are self-employed ans those employed on wages take. The plot describes that there is no such distinction between the sleep hours. Self-employed and employed on wages people both take nearly similar hours of sleep. The median is found to be 7 hours when we visually analyze the graph. 







**Research quesion 3:**

```{r}
health <- brfss2013%>%
  select(income2,menthlth)%>%
  filter(!is.na(menthlth) & income2 == "$75,000 or more")%>%
  group_by(income2,menthlth)%>%
  summarize(count=n())
health
```

The summary statistics shows the count of people with number of poor mental health days in a month who earns more than 75000 dollars anually. 

```{r}
ggplot(health, aes(x=menthlth,y=count,fill = income2))+geom_col()
                                          
```

The plot shows the distribution of the number of poor mental health days in a month for people who earns more than  75000 dollars annually. We can interpret that the most of the people dont suffer from any mental health issues, but there are few outliers who face more than 10 days of poor mental health. 






